---
title: "CPLN 5920 Final Project: EDA"
format: html
editor: visual
---

## Start with the outcome variable (filings)
1. How skewed is it?
2. How many zeros?
3. Overdispersion?

```{r setup, include=FALSE}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidycensus)


# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

```{r load data set}
evict <- read_csv("data/eviction.csv")

glimpse(evict)

evict <- evict %>%
  mutate(
    GEOID = as.character(GEOID),
    month_date = my(month),   # "03/2023" → 2023-03-01
    year = year(month_date),
    month_num = month(month_date)
  ) %>%
  dplyr::select(-last_updated)

```

---
## Distribution

```{r}
ggplot(evict, aes(x = filings_2020)) +
  geom_histogram(bins = 50)

```
Eviction filings are highly skewed and there are a lot of zeros.

---
## Time Trends
1. Upward/downward trends
2. Seasonal patterns

```{r}
evict %>%
  group_by(month_date) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE)) %>%
  ggplot(aes(x = month_date, y = avg_filings)) +
  geom_line()

```
There is no strong long-term trend in the filing data and very high month-to-month volatility.
---
## Spatial Inequality
1. Top & bottom tracts
2. top 10 tracts by absolute filing numbers and concentration

```{r}
evict %>%
  group_by(GEOID) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE)) %>%
  ggplot(aes(x = avg_filings)) +
  geom_histogram(bins = 50)

```

```{r}
top10 <- evict %>%
  group_by(GEOID) %>%
  summarise(total = sum(filings_2020, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  head(10) # Top 10 census tracts with the most filings

top10

# Pull renter households data
renters <- get_acs(
  geography = "tract",
  variables = "B25003_003",
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  geometry = FALSE
)

renters <- renters %>%
  dplyr::select(GEOID, renters = estimate)

# Join to dataset
evict <- evict %>%
  left_join(renters, by = "GEOID")

# Normaliza filings
evict <- evict %>%
  mutate(
    filings_per_1k_renters = (filings_2020 / renters) * 1000
  )

# Compute tract-level average rate
evict_clean_rate <- evict %>%
  filter(!is.na(renters) & renters > 0)

evict_clean_rate <- evict_clean_rate %>%
  filter(renters > 100)   # adjust threshold

top10_rate <- evict_clean_rate %>%
  group_by(GEOID) %>%
  summarise(
    avg_rate = mean(filings_per_1k_renters, na.rm = TRUE),
    total_filings = sum(filings_2020, na.rm = TRUE),
    renters = first(renters),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_rate)) %>%
  slice(1:10)

top10_rate

# Find common tracts as policy hotspots
hotspot <- intersect(top10$GEOID, top10_rate$GEOID)
hotspot

```

There are a few tracts that have very high filing records. By finding the top 10 census tracts with highest filing records and top 10 census tracts with highest filing record concentration, we identify two high risk census tracts that appear in both lists as rental assistance policy hotspots.
---
## Dynamics
1. Create multiple lag variables (1 month lag, 1 quarter lag)
2. Visualize Changes

```{r}
evict <- evict %>%
  arrange(GEOID, month_date) %>%
  group_by(GEOID)

# 1 month lag
evict <- evict %>%
  mutate(
    lag_1m = lag(filings_2020, 1)
  )

# Quarter lag
evict <- evict %>%
  mutate(
    lag_3m = lag(filings_2020, 3)
  )

# Create change variables
evict <- evict %>%
  mutate(
    change_1m = filings_2020 - lag_1m,
    change_3m = filings_2020 - lag_3m
  )

# Visualize 1 month change
ggplot(evict, aes(x = change_1m)) +
  geom_histogram(bins = 50) +
  labs(title = "Monthly Change in Filings")

# Visualize quarterly change
ggplot(evict, aes(x = change_3m)) +
  geom_histogram(bins = 50) +
  labs(title = "Quarterly Change in Filings")

```
Both monthly and quarterly change have symmetrical distribution. Eviction filings are not consistently increasing or decreasing. Changes in eviction activity fluctuate around a stable baseline rather than showing persistent growth or decline. This suggests that recent lag components of filings are strong predictors for future risks.

---
## Race/Demographic Patterns

```{r}
evict %>%
  group_by(racial_majority) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE))

```
Black-majority neighborhoods have substantially higher average eviction filings. Even after adjusting for renter population, eviction activity remains higher in Black-majority neighborhoods, suggesting structural disparities in housing stability.
---






























