---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Gab Chen"
date: 2026-01-29
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Philadelphia Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)

# Set your Census API key
census_api_key("e2510a23b0c3dba6ac091ea554e6f2605dbde651", install = TRUE, overwrite = TRUE)

# Choose your state for analysis - assign it to a variable called my_state
```

**State Selection:** I have chosen **Pennsylvania** for this analysis because: For consistency.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
pa <- get_acs(geography = "county", variables = c(totpop = "B01003_001", medinc = "B19013_001"),
        state = "PA", survey = "acs5", year = 2022, output = "wide")

# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
pa_clean <- pa %>%
  mutate(
    county_name = str_remove(NAME, " County, Pennsylvania")
  )

# Display the first few rows
head(pa_clean)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
pa_reliability <- pa_clean %>%
  mutate(
    moe_percentage = round((medincM / medincE) * 100, 2),
    reliability = case_when(
      moe_percentage < 5 ~ "High Confidence",
      moe_percentage >= 5 & moe_percentage <= 10 ~ "Moderate",
      moe_percentage > 10 ~ "Low Confidence"
    )
  )

# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
pa_reliability %>%
  count(reliability) %>%
  mutate(
    percent = round(n / sum(n) * 100, 2)
  )

```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 3 counties by MOE percentage
pa_reliability %>%
  arrange(desc(moe_percentage)) %>%
  slice(1:5) %>%
  select(county_name, medincE, moe_percentage, reliability) %>%
# Format as table with kable() - include appropriate column names and caption
kable(col.names = c(
  "County", 
  "Median Household Income", 
  "MOE Percentage (%)", 
  "Reliability Category"), 
  caption = "Top 5 Pennsylvania Counties with Highest Income Estimate Uncertainty"
)

```

**Data Quality Commentary:**

[Counties such as Forest, Sullivan, and Union show relatively high uncertainty in median household income estimates, which means algorithms relying on this data may make less reliable decisions for these areas. This could lead to misallocation of resources or misclassification in models used for funding, eligibility, or planning. Higher uncertainty in these counties is likely driven by small populations, rural characteristics, and limited survey samples, which increase margins of error in ACS estimates.]


# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_counties <- pa_reliability %>%
  filter(moe_percentage > 9.5 |
  moe_percentage > 4.8 & moe_percentage < 5)

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
print(selected_counties) %>%
select(county_name, medincE, moe_percentage, reliability)
```

**Comment on the output:** [Using filter() to select a certain number of observations requires you to have a basic understanding of your data frame.One county with the highest MOE percentage is selected from each reliability group.]

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
race_vars <- c(
  white = "B03002_003",  # white alone
  black = "B03002_004", # Black/African American
  his_lat = "B03002_012", # Hispanic/Latino
  totpop = "B03002_001" # total population
)
# Use get_acs() to retrieve tract-level data
tract_data <- get_acs(
  geography = "tract",
  variables = race_vars,
  state = "PA",
  county = c("42053","42103"),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
# Hint: You may need to specify county codes in the county parameter

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
pct_race <- tract_data %>%
  mutate(
    pct_white = round(whiteE / totpopE, 3) * 100,
    pct_black = round(blackE / totpopE, 3) * 100,
    pct_his_lat = round(his_latE / totpopE, 3) * 100
  )

# Add readable tract and county name columns using str_extract() or similar
pct_race_clean <- pct_race %>%
  mutate(
    TRACT = str_extract(NAME, "\\d+.?\\d+" ),
    COUNTY = str_extract(NAME, "(?<=;)\\s[^ ]+")
    )
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
pct_race_clean %>%
  arrange(desc(pct_his_lat)) %>%
  slice(1)

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
pct_race_clean %>%
  group_by(COUNTY) %>%
  summarize(
    n_tracts = n(),
    pct_white = round(mean(pct_white, na.rm = TRUE), 1),
    pct_black = round(mean(pct_black, na.rm = TRUE), 1),
    pct_his_lat = round(mean(pct_his_lat, na.rm = TRUE), 1)
  ) %>%

# Create a nicely formatted table of your results using kable()
kable(col.names = c(
  "County", 
  "Number of Tracts", 
  "Percent of White Residents (%)", 
  "Percent of Black Residents (%)", 
  "Percent of Hispanic or Latino Residents (%)"
  ),
  caption = "Average Demographis by County"
  )

```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
pct_race_clean <- pct_race_clean %>%
  mutate(
    MOE_white = round(whiteM / whiteE * 100, 2),
    MOE_black = round(blackM / blackE * 100, 2),
    MOE_his_lat = round(his_latM / his_latE * 100, 2)
  ) 

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
flag_tract <- pct_race_clean %>%
  mutate(
    MOE_issue = if_else(
      MOE_white > 15 | MOE_black > 15 | MOE_his_lat > 15,
      TRUE,
      FALSE
    )
  )

# Create summary statistics showing how many tracts have data quality issues
flag_tract %>%
  summarise(
    tracts_with_MOE_issues = sum(MOE_issue, na.rm = TRUE)
  )
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Categorize MOE > 100 as "high", MOE <= 100 as "low"
flag_tract <- flag_tract %>%
  mutate(
    MOE_category = case_when(
      MOE_white > 100 | MOE_black > 100 | MOE_his_lat > 100 ~ "high",
      MOE_white <= 100 | MOE_black <= 100 | MOE_his_lat <= 100 ~ "low",
    )
  )

# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison
flag_tract %>%
group_by(MOE_category) %>%
summarise(
  avg_pop = round(mean(totpopE, na.rm = TRUE), 0),
  avg_pct_white = round(mean(pct_white, na.rm = TRUE), 1),
  avg_pct_black = round(mean(pct_black, na.rm = TRUE), 1),
  avg_pct_hispanic = round(mean(pct_his_lat, na.rm = TRUE), 1)
) %>%

# Create a professional table showing the patterns
kable(
  col.names = c(
    "MOE Issue Category",
    "Average Population Size",
    "Average percentage of white population (%)",
    "Average percentage of black population (%)",
    "Average percentage of hispanic or latino population (%)"
  ),
  caption = "Average Demographic Characteristics for Tracts with Different Levels of MOE Issues"
)

```

**Pattern Analysis:** [Tracts with high MOE issues have a smaller population size and more demographically homogeneous, with a larger share of white presence and smaller share of black and hispanic or latino population. Tracts with smaller population size have smaller survey samples. More homogeneous tracts have limited variation across demographic categories. Both of the cases increase statistical uncertainty and inflate margins of error.]


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

[Your integrated 4-paragraph summary here]

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** [List counties with high confidence data and explain why they're appropriate]

2. **Counties requiring additional oversight:** [List counties with moderate confidence data and describe what kind of monitoring would be needed]

3. **Counties needing alternative approaches:** [List counties with low confidence data and suggest specific alternatives - manual review, additional surveys, etc.]

## Questions for Further Investigation

[List 2-3 questions that your analysis raised that you'd like to explore further in future assignments. Consider questions about spatial patterns, time trends, or other demographic factors.]

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [date]

**Reproducibility:** 
- All analysis conducted in R version [your version]
- Census API key required for replication
- Complete code and documentation available at: [your portfolio URL]

**Methodology Notes:**
[Describe any decisions you made about data processing, county selection, or analytical choices that might affect reproducibility]

**Limitations:**
[Note any limitations in your analysis - sample size issues, geographic scope, temporal factors, etc.]

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`